---
title: "What It Takes To Win: Cricket - Women's T20"
author: "Jacquie Tran | `r Sys.Date()`"
format:
  revealjs:
    theme: [simple, custom.scss]
    incremental: true
    chalkboard: true
---

```{r setup}

# Define document colours ------------------------------------------------------

colours_light_grey <- "#CAC1BE"
colours_amaranth <- "#9D1B39"
colours_aquamarine <- "#15F4B5"
colours_mint <- "#09C38E"

```


## Background

- Cricket will be added to the Olympic program for LA 2028. The format is expected to be T20.
- The Australian Women's cricket team has long been among the most successful international sides in world cricket.
- With the professionalisation of women's cricket growing globally, more nations are challenging the Australian Women's team in major events **(see: T20 World Cup 2024, won by New Zealand)**

## Key rules of T20 cricket

- Two teams of 11 players each. Teams take turns to bat and bowl.
- In the T20 format, teams will complete one innings of each mode.
- A standard innings is made up of 20 overs, with each over consisting of 6 regulation balls bowled, i.e., 120 regulation balls in total.
     - Non-regulation balls are called "extras" and concede runs when bowled.
- An innings ends when all regulation balls have been bowled, or when the batting side has lost all 10 wickets (batting must be done in partnerships).
- A match result is achieved when:
     - The team that bats 1st sets a target that the team batting 2nd does not chase down before the end of their innings, or
     - The team that bats 2nd successfully chases the target set by the team that batted 1st, with at least 1 wicket still in hand.

## Bottom Line Up Front

From this exploratory analysis of game data from T20 World Cup tournament play, these factors appear related to winning in Women's T20 cricket:

**Batting**

- Point 1
- Point 2
- Point 3

**Bowling**

- Point 1
- Point 2
- Point 3

# Methods

## Caveat: Only a slice of "What It Takes To Win"

This quantitative analysis uses publicly available competition and game data only

Other factors are critical to performance but not reflected in public data, and would benefit from mixed methods analysis, for example:

- Daily training measures
- Measures of training state and performance: Physiological, biomechanical, physical, psychological
- "Tight team": Trusted relationships between athletes, coaches, and support team

## Overview of analytical approach

**Game outcomes**

- Wickets
- Runs scored

**Performance measures**

- Scoring: *"Batting performance characteristics appear to be more important than bowling characteristics, which is even more pronounced in women's T20 cricket."* ([Bhardwaj & Dwyer, 2022](https://doi.org/10.1080/24748668.2022.2045820))

- "Tempo": Bowlers increasing pressure on the batting team by challenging finite resources (balls and wickets)

*Note: The 3rd part of the game, fielding, has not been included in this analysis, due to limited fielding-specific measures available. Fielding is part of a bowling innings.*

## Data source

```{r retrieve_data, cache = TRUE, message = FALSE}

# Obtain match IDs for Women's T20 World Cup tournaments or the 2022
# Commonwealth Games
t20wc_match_ids <- cricketdata::fetch_cricsheet(
  type = "match",
  gender = "female",
  competition = "t20s") |>
  # Filter to World Cup matches
  dplyr::filter(
    event %in% c(
      "ICC Women's T20 World Cup",
      "Commonwealth Games Women's Cricket Competition"))

# Retrieve ball-by-ball data, filtering to keep only matches from T20 World Cups
# or the 2022 Commonwealth Games
t20wc_bbb <- cricketdata::fetch_cricsheet(
  type = "bbb",
  gender = "female",
  competition = "t20s") |>
  # Filter to World Cup matches
  dplyr::filter(
    match_id %in% t20wc_match_ids$match_id)

```

```{r tidy_data}

# Tidy match IDs data frame
t20wc_match_ids_tidy <- t20wc_match_ids |>
  dplyr::mutate(
    match_id = as.numeric(match_id)) |>
  dplyr::select(
    match_id, event, winner, winner_runs, winner_wickets)

# Merge data frames together and tidy ball-by-ball data
t20wc_bbb_tidy <- dplyr::left_join(
  t20wc_bbb, t20wc_match_ids_tidy, by = "match_id") |>
  dplyr::mutate(
    year = stringr::str_sub(start_date, start = 1, end = 4)) |>
  dplyr::select(
    match_id, season, year, event, start_date, dplyr::starts_with("winner"),
    innings, over, ball, dplyr::ends_with("_team"), dplyr::contains("striker"),
    bowler, runs_off_bat, extras, ball_in_over, extra_ball, balls_remaining,
    runs_scored_yet, wicket, wickets_lost_yet, dplyr::ends_with("_total"),
    wides, noballs, byes, legbyes, penalty, wicket_type, player_dismissed,
    other_wicket_type, other_player_dismissed)

```

```{r calculate_sample_characteristics}

# Count all matches
count_all_matches <- t20wc_match_ids_tidy |>
  dplyr::summarise(n = dplyr::n()) |>
  dplyr::pull(n)

# Count the number of matches with a result
count_matches_with_result <- t20wc_match_ids_tidy |>
  dplyr::filter(!is.na(winner)) |>
  dplyr::summarise(n = dplyr::n()) |>
  dplyr::pull(n)

# Count the number of teams represented
count_number_of_teams <- t20wc_bbb_tidy |>
  dplyr::distinct(batting_team) |>
  dplyr::summarise(n = dplyr::n()) |>
  dplyr::pull(n)

```

This analysis draws on ball-by-ball data from international women's T20 tournaments from 2018 to 2024, specifically including:

- Women's T20 World Cup tournaments held in 2018, 2020, 2023, and 2024
- The Birmingham 2022 Commonwealth Games
- Time span reflects the extent of publicly available ball-by-ball data
     
Data obtained from [Cricsheet.org](https://cricsheet.org) using the `{cricketdata}` R package:

- `r count_matches_with_result` matches with a result (`r count_all_matches - count_matches_with_result` match excluded from analysis of performance indicators due to no result)
- `r count_number_of_teams` teams across the 5 tournaments in the data set

# Game outcomes

High-level performance outcomes

## Ranking teams on match outcomes

Ranked by tournament wins, then win percentage, then losses (ascending)

```{r wrangle_team_rankings}

# Count number of tournaments played by each team
teams_tournament_count <- t20wc_match_ids |>
  dplyr::mutate(
    year = stringr::str_sub(date, start = 1, end = 4)) |>
  dplyr::select(
    year, match_id, team1, team2) |>
  tidyr::pivot_longer(
    cols = c("team1", "team2"),
    names_to = "key",
    values_to = "team") |>
  dplyr::select(
    year, team) |>
  dplyr::group_by(year) |>
  dplyr::distinct(team, .keep_all = TRUE) |>
  dplyr::ungroup() |>
  dplyr::group_by(team) |>
  dplyr::summarise(
    n_tournaments = dplyr::n(),
    .groups = "drop")

# Summarise match results for each team
team_match_outcomes <- t20wc_match_ids |>
  dplyr::select(
    match_id, team1, team2, winner) |>
  tidyr::pivot_longer(
    cols = c("team1", "team2"),
    names_to = "key",
    values_to = "team") |>
  dplyr::mutate(
    result = dplyr::case_when(
      is.na(winner)  ~ "no_result",
      winner == team ~ "win",
      winner != team ~ "loss")) |>
  dplyr::select(
    match_id, team, result) |>
  dplyr::group_by(team, result) |>
  dplyr::summarise(
    n_matches = dplyr::n(),
    .groups = "drop") |>
  tidyr::pivot_wider(
    id_cols = "team",
    names_from = "result",
    values_from = "n_matches",
    values_fill = 0) |>
  dplyr::mutate(
    matches_total = win + loss + no_result,
    win_pct = round(win / matches_total * 100, 1)) |>
  dplyr::select(
    team, win, loss, no_result, matches_total, win_pct)

# Merge tournaments count with match outcomes
team_outcomes_tidy <- dplyr::left_join(
  teams_tournament_count, team_match_outcomes,
  by = "team") |>
  dplyr::arrange(
    desc(win), desc(win_pct), loss) |>
  # Add emojis to denote trophies won, finalists, and medallists
  dplyr::mutate(
    team_label = dplyr::case_when(
      team == "Australia"    ~ paste0(
        team, " ", emoji::emoji("trophy"), emoji::emoji("trophy"),
        emoji::emoji("trophy"), emoji::emoji("1st_place_medal")),
      team == "New Zealand"  ~ paste0(
        team, " ", emoji::emoji("trophy"), emoji::emoji("3rd_place_medal")),
      team == "India"        ~ paste0(
        team, " ", emoji::emoji("chequered_flag"), emoji::emoji("2nd_place_medal")),
      team == "England"      ~ paste0(
        team, " ", emoji::emoji("chequered_flag")),
      team == "South Africa" ~ paste0(
        team, " ", emoji::emoji("chequered_flag"),
        emoji::emoji("chequered_flag")),
      TRUE                ~ team)) |>
  dplyr::relocate(team_label)

```

```{r tabulate_team_outcomes_tidy}

team_outcomes_tidy |>
  dplyr::select(-team) |>
  gt::gt() |>
  gt::cols_label(
    team_label = "",
    n_tournaments = "# of tournaments",
    win = "Wins",
    loss = "Losses",
    no_result = "No result",
    matches_total = "Total matches",
    win_pct = "Win %") |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()) |>
  gt::data_color(
    columns = win,
    method = "numeric",
    palette = c(colours_amaranth, colours_aquamarine),
    domain = c(
      min(team_outcomes_tidy$win), max(team_outcomes_tidy$win))) |>
  gt::data_color(
    columns = win_pct,
    method = "numeric",
    palette = c(colours_amaranth, colours_aquamarine),
    domain = c(
      min(team_outcomes_tidy$win_pct), max(team_outcomes_tidy$win_pct))) |>
  gt::data_color(
    columns = loss,
    method = "numeric",
    palette = c(colours_aquamarine, colours_amaranth),
    domain = c(
      min(team_outcomes_tidy$loss), max(team_outcomes_tidy$loss)))

```



## Ranking teams on wins - superceded

Rank on number of matches won, then win percentage

```{r wrangle_teams_in_3_plus_tournaments}

# First, identify teams that played in 3+ tournaments
teams_in_3_plus_tournaments <- t20wc_match_ids |>
  dplyr::select(
    date, team1, team2) |>
  dplyr::mutate(
    year = stringr::str_sub(date, start = 1, end = 4)) |>
  dplyr::select(
    year, team1, team2) |>
  tidyr::pivot_longer(
    cols = c("team1", "team2"),
    names_to = "key",
    values_to = "value") |>
  dplyr::select(-key) |>
  dplyr::group_by(year) |>
  dplyr::distinct(value, .keep_all = TRUE) |>
  dplyr::ungroup() |>
  dplyr::group_by(value) |>
  dplyr::summarise(
    n_tournaments = dplyr::n(),
    .groups = "drop") |>
  dplyr::filter(n_tournaments >= 3)

```


```{r wrangle_team_rankings_superceded, eval = FALSE}

team_rankings <- t20wc_match_ids |>
  dplyr::select(
    match_id, team1, team2, winner) |>
  tidyr::pivot_longer(
    cols = c("team1", "team2"),
    names_to = "key",
    values_to = "team") |>
  dplyr::mutate(
    result = dplyr::case_when(
      is.na(winner)  ~ "no_result",
      winner == team ~ "win",
      winner != team ~ "loss")) |>
  # Filter to teams in 3+ tournaments
  dplyr::filter(
    team %in% teams_in_3_plus_tournaments$value) |>
  dplyr::select(
    match_id, team, result) |>
  dplyr::group_by(team, result) |>
  dplyr::summarise(
    n_matches = dplyr::n(),
    .groups = "drop") |>
  tidyr::pivot_wider(
    id_cols = "team",
    names_from = "result",
    values_from = "n_matches",
    values_fill = 0) |>
  dplyr::mutate(
    matches_total = win + loss + no_result,
    win_pct = round(win / matches_total * 100, 1)) |>
  dplyr::arrange(
    desc(win), desc(win_pct)) |>
  dplyr::select(
    team, win, loss, no_result, matches_total, win_pct)

```

```{r tabulate_team_rankings, eval = FALSE}

team_rankings |>
  gt::gt()

```

## Wickets taken in wins vs. losses

```{r wrangle_wickets_summary}

# Isolate wickets taken from ball-by-ball data
wickets_summary <- t20wc_bbb_tidy |>
  dplyr::filter(!is.na(winner)) |> # Exclude matches with no results
  dplyr::filter(wicket == TRUE) |>
  dplyr::select(
    year, match_id, winner, bowling_team, batting_team, innings, wicket) |>
  dplyr::group_by(
    year, match_id, winner, innings, bowling_team) |>
  dplyr::summarise(
    n_wickets = length(wicket),
    .groups = "drop") |>
  dplyr::mutate(
    flag_bowling_team_result = dplyr::case_when(
      winner == bowling_team ~ 1,
      winner != bowling_team ~ 0),
    bowling_team_result_label = dplyr::case_when(
      flag_bowling_team_result == 1 ~ "Win",
      flag_bowling_team_result == 0 ~ "Loss"),
    innings_label = dplyr::case_when(
      innings == 1 ~ "Bowling 1st",
      innings == 2 ~ "Bowling 2nd"))

# Calculate median wickets taken in wins and losses, when bowling 1st and 2nd
wickets_medians <- wickets_summary |>
  dplyr::group_by(innings_label, flag_bowling_team_result) |>
  dplyr::summarise(
    n_wickets = median(n_wickets),
    .groups = "drop")

```

```{r plot_wickets_summary, fig.height = 3.5, message = FALSE}

# Build plot
plot_wickets_summary <- ggplot2::ggplot() +
  ggplot2::geom_linerange(
    data = wickets_medians,
    ggplot2::aes(
      x = n_wickets, ymin = flag_bowling_team_result + 0.5,
      ymax = flag_bowling_team_result + 1.5),
    linewidth = 1.2) +
  ggbeeswarm::geom_quasirandom(
    data = wickets_summary,
    ggplot2::aes(
      x = n_wickets, y = factor(flag_bowling_team_result),
      group = flag_bowling_team_result,
      colour = factor(flag_bowling_team_result)),
    groupOnX = FALSE, size = 3.5, alpha = 0.7, shape = 19, stroke = NA) +
  ggplot2::facet_wrap(~innings_label) +
  ggplot2::scale_x_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, by = 2)) +
  ggplot2::scale_y_discrete(
    labels = c(
      "1" = "Win",
      "0" = "Loss")) +
  ggplot2::scale_colour_manual(
    values = c(
      "1" = colours_mint,
      "0" = colours_amaranth)) +
  ggplot2::labs(
    x = "Wickets taken", y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor.x = ggplot2::element_blank(),
    panel.spacing = grid::unit(2, "lines"),
    strip.text = ggplot2::element_text(
      size = ggplot2::rel(1.4), face = "bold"),
    strip.background = ggplot2::element_rect(
      colour = colours_light_grey, fill = colours_light_grey),
    axis.title.x = ggplot2::element_text(
      size = ggplot2::rel(1.6), margin = ggplot2::margin(t = 10, unit = "pt")),
    axis.text.x = ggplot2::element_text(
      size = ggplot2::rel(1.6)),
    axis.text.y = ggplot2::element_text(
      size = ggplot2::rel(1.6), face = "bold")
  )

# Display plot
plot_wickets_summary

```

## Another way to consider wickets

```{r wrangle_wickets_and_win_pct}

# Determine % games won / lost based on categorical grouping of wickets taken
wickets_and_win_pct <- wickets_summary |>
  dplyr::mutate(
    wickets_category = dplyr::case_when(
      innings == 1 & n_wickets >= 7 ~ "7+ wickets",
      innings == 1 & n_wickets < 7  ~ "< 7 wickets",
      innings == 2 & n_wickets >= 5 ~ "5+ wickets",
      innings == 2 & n_wickets < 5  ~ "< 5 wickets")) |>
  dplyr::group_by(
    innings_label, wickets_category, bowling_team_result_label) |>
  dplyr::summarise(
    n_matches = dplyr::n(),
    .groups = "drop") |>
  dplyr::group_by(
    innings_label, wickets_category) |>
  dplyr::mutate(
    subtotal_matches = sum(n_matches)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    matches_as_pct_subtotal = round(
      n_matches / subtotal_matches * 100, 0),
    n_label = paste0("n = ", n_matches),
    pct_label = paste0(matches_as_pct_subtotal, "%"),
    combined_label = dplyr::case_when(
      wickets_category == "< 5 wickets" &
        bowling_team_result_label == "Win" ~ paste0(
          pct_label, ", ", n_label),
      TRUE                                 ~ paste0(
        pct_label, "\n", n_label))) |>
  dplyr::group_by(wickets_category) |>
  dplyr::arrange(
    wickets_category, desc(bowling_team_result_label)) |>
  dplyr::mutate(
    lab_xpos = cumsum(matches_as_pct_subtotal) - 0.5 * matches_as_pct_subtotal) |>
  dplyr::ungroup()

```

```{r plot_wrangle_wickets_and_win_pct, fig.height = 3.5}

# Build plot
plot_wrangle_wickets_and_win_pct <- ggplot2::ggplot() +
  ggplot2::geom_bar(
    data = wickets_and_win_pct,
    ggplot2::aes(
      x = matches_as_pct_subtotal, y = wickets_category,
      fill = bowling_team_result_label),
    position = "stack", stat = "identity", width = 0.7) +
  ggplot2::geom_text(
    data = wickets_and_win_pct |>
      dplyr::filter(
        bowling_team_result_label == "Win" &
          wickets_category != "< 5 wickets"),
    ggplot2::aes(
      x = lab_xpos, y = wickets_category, label = combined_label),
    size = 4, fontface = "bold", lineheight = 0.9) +
  ggplot2::geom_text(
    data = wickets_and_win_pct |>
      dplyr::filter(
        bowling_team_result_label == "Win" &
          wickets_category == "< 5 wickets"),
    ggplot2::aes(
      x = lab_xpos, y = wickets_category, label = combined_label),
    size = 4, fontface = "bold", lineheight = 0.9,
    nudge_x = 8, nudge_y = 0.45) +
  ggplot2::geom_text(
    data = wickets_and_win_pct |>
      dplyr::filter(
        bowling_team_result_label == "Loss" &
          wickets_category != "5+ wickets"),
    ggplot2::aes(
      x = lab_xpos, y = wickets_category, label = combined_label),
    size = 4, fontface = "bold", lineheight = 0.9, colour = "#FFFFFF") +
  ggplot2::geom_text(
    data = wickets_and_win_pct |>
      dplyr::filter(
        bowling_team_result_label == "Loss" &
          wickets_category == "5+ wickets"),
    ggplot2::aes(
      x = lab_xpos, y = wickets_category, label = combined_label),
    size = 3, fontface = "bold", lineheight = 0.9, colour = "#FFFFFF") +
  ggplot2::facet_wrap(~innings_label, scales = "free_y") +
  ggplot2::scale_fill_manual(
    values = c(
      "Win"  = colours_mint,
      "Loss" = colours_amaranth)) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "top",
    legend.title = ggplot2::element_blank(),
    legend.text = ggplot2::element_text(
      size = ggplot2::rel(1.2)),
    legend.key.spacing.x = grid::unit(15, "pt"),
    panel.grid.major = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    panel.spacing = grid::unit(2, "lines"),
    strip.text = ggplot2::element_text(
      size = ggplot2::rel(1.4), face = "bold"),
    strip.background = ggplot2::element_rect(
      colour = colours_light_grey, fill = colours_light_grey),
    axis.title.x = ggplot2::element_text(
      size = ggplot2::rel(1.6), margin = ggplot2::margin(t = 10, unit = "pt")),
    axis.text.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(
      size = ggplot2::rel(1.6), face = "bold")) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(reverse = TRUE))

# Display plot
plot_wrangle_wickets_and_win_pct

```

Teams generally went on to win the game when...

- Bowling 1st and taking 7+ wickets
- Bowling 2nd and taking 5+ wickets
- These bowling stats also seem to reflect how batting teams approach an innings when setting (aggressive) vs. chasing (conservative)

## Wickets in phases

```{r wrangle_wickets_in_phases}

wickets_in_phases <- t20wc_bbb_tidy |>
  dplyr::filter(!is.na(winner)) |> # Exclude matches with no results
  dplyr::filter(wicket == TRUE) |>
  dplyr::select(
    year, match_id, winner, bowling_team, batting_team, innings, over, wicket) |>
  dplyr::mutate(
    phase_of_innings = dplyr::case_when(
      over >= 1 & over <= 6  ~ "Powerplay",
      over >= 7 & over <= 15 ~ "Consolidation phase",
      over >= 16             ~ "Death overs")) |>
  dplyr::group_by(
    year, match_id, winner, bowling_team, phase_of_innings) |>
  dplyr::summarise(
    n_wickets = length(wicket),
    .groups = "drop") |>
  dplyr::mutate(
    flag_bowling_team_result = dplyr::case_when(
      winner == bowling_team ~ 1,
      winner != bowling_team ~ 0),
    bowling_team_result_label = dplyr::case_when(
      flag_bowling_team_result == 1 ~ "Win",
      flag_bowling_team_result == 0 ~ "Loss"),
    phase_of_innings = factor(
      phase_of_innings,
      levels = c("Powerplay", "Consolidation phase", "Death overs")),
    bowling_team_result_label = factor(
      bowling_team_result_label,
      levels = c("Win", "Loss")))

```

```{r plot_wickets_in_phases}

# Build plot
plot_wickets_in_phases <- ggplot2::ggplot() +
  ggplot2::geom_boxplot(
    data = wickets_in_phases,
    ggplot2::aes(
      x = phase_of_innings, y = n_wickets, group = phase_of_innings,
      colour = bowling_team_result_label, fill = bowling_team_result_label),
    outlier.shape = NA, width = 0.7, alpha = 0.2, linewidth = 1.25,
    fatten = 2.5) +
  ggplot2::facet_wrap(~bowling_team_result_label) +
  ggplot2::scale_x_discrete(
    labels = c(
      c("Powerplay\n(overs 1-6)",
        "Consolidation\n(overs 7-15)",
        "Death overs\n(overs 16-20)"))) +
  ggplot2::scale_colour_manual(
    values = c(
      "Win" = colours_mint,
      "Loss" = colours_amaranth)) +
  ggplot2::scale_fill_manual(
    values = c(
      "Win" = colours_mint,
      "Loss" = colours_amaranth)) +
  ggplot2::labs(
    x = NULL, y = "Number of wickets taken") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    panel.spacing = grid::unit(2, "lines"),
    strip.text = ggplot2::element_text(
      size = ggplot2::rel(1.4), face = "bold"),
    strip.background = ggplot2::element_rect(
      colour = colours_light_grey, fill = colours_light_grey),
    axis.title.y = ggplot2::element_text(
      size = ggplot2::rel(1.6), margin = ggplot2::margin(r = 10, unit = "pt")),
    axis.text.x = ggplot2::element_text(
      size = ggplot2::rel(1.4)),
    axis.text.y = ggplot2::element_text(
      size = ggplot2::rel(1.6))
  )

# Display plot
plot_wickets_in_phases

```


## Runs: Innings totals

```{r wrangle_innings_totals}

innings_totals <- t20wc_bbb_tidy |>
  dplyr::select(
    match_id, year, winner, innings, batting_team, runs_scored_yet) |>
  dplyr::group_by(
    match_id, year, winner, innings, batting_team) |>
  dplyr::summarise(
    innings_total = max(runs_scored_yet),
    .groups = "drop") |>
  dplyr::mutate(
    result_for_batting_team = dplyr::case_when(
      winner == batting_team ~ 1,
      TRUE                   ~ 0)) |>
  # Filter out matches with no result
  dplyr::filter(match_id != 1173066) |>
  dplyr::mutate(
    innings_label = dplyr::case_when(
      innings == 1 ~ "Setting",
      innings == 2 ~ "Chasing"),
    result_label = dplyr::case_when(
      result_for_batting_team == 1 ~ "Win",
      result_for_batting_team == 0 ~ "Loss"),
    innings_label = factor(
      innings_label,
      levels = c("Setting", "Chasing")),
    result_label = factor(
      result_label,
      levels = c("Win", "Loss"))
    )

```

- Sample: `r length(innings_totals$match_id)` matches from `r length(unique(t20wc_bbb_tidy$year))` tournaments
- 1 match excluded due to no result (rained out)

```{r analyse_innings_totals}

# Calculate median winning and losing innings totals,
# when chasing and setting
innings_total_summary <- innings_totals |>
  dplyr::group_by(
    year, innings_label, result_label) |>
  dplyr::summarise(
    innings_total_median = median(innings_total),
    .groups = "drop"
  )

```

```{r plot_innings_totals_pooled, message = FALSE}

ggplot2::ggplot() +
  ggplot2::geom_violin(
    data = innings_totals,
    ggplot2::aes(
      x = innings_total, y = result_for_batting_team,
      group = result_for_batting_team)) +
  ggbeeswarm::geom_quasirandom(
    data = innings_totals,
    ggplot2::aes(
      x = innings_total, y = result_for_batting_team,
      group = result_for_batting_team),
    groupOnX = FALSE, varwidth = TRUE,
    method = "pseudorandom", cex = 2.5) +
  ggplot2::facet_wrap(~innings_label)

```
 
- Wide variability in totals when setting, but ~150 runs seems important.
 - No discernible difference in chasing totals between winning and losing teams (To be expected - because the winning target is set by the team that bats first)
 
## Setting a total above / below 150
 
```{r analyse_innings_above_below_150}

# Focus on setting innings only
innings_above_below_150 <- innings_totals |>
  dplyr::filter(innings == 1) |>
  dplyr::mutate(
    flag_above_below_150 = dplyr::case_when(
      innings_total >= 150 ~ "150+ runs",
      innings_total < 150  ~ "< 150 runs")) |>
  dplyr::group_by(
    result_label, flag_above_below_150) |>
  dplyr::summarise(
    n_innings = dplyr::n(),
    .groups = "drop")

```

```{r plot_innings_above_below_150, message = FALSE}

# Build plot
plot_innings_above_below_150 <- ggplot2::ggplot() +
  ggbeeswarm::geom_quasirandom(
    data = innings_totals |>
      dplyr::filter(innings_label == "Setting"),
    ggplot2::aes(
      x = innings_total, y = result_for_batting_team,
      group = result_for_batting_team, colour = innings_total >= 150),
    groupOnX = FALSE, varwidth = TRUE,
    method = "pseudorandom", cex = 2.5)

# Display plot
plot_innings_above_below_150

```


The team batting first has won:

- 97% of matches when setting a target of 150 or above
- 37% of matches when setting a target below 150

# Batting performance

## Boundary scoring rate

Note: Cricsheet data does not flag boundary-scoring balls, as such. For this analysis, I am working under the assumption that "runs off bat" of 4 or 6 are boundaries. It is technically possible to run 4 or 6 runs off one ball (say, with overthrows), but these are rare occurrences.

```{r wrangle_boundary_scoring}

# First, identify teams that played in 3+ tournaments
teams_in_3_plus_tournaments <- t20wc_match_ids |>
  dplyr::select(
    date, team1, team2) |>
  dplyr::mutate(
    year = stringr::str_sub(date, start = 1, end = 4)) |>
  dplyr::select(
    year, team1, team2) |>
  tidyr::pivot_longer(
    cols = c("team1", "team2"),
    names_to = "key",
    values_to = "value") |>
  dplyr::select(-key) |>
  dplyr::group_by(year) |>
  dplyr::distinct(value, .keep_all = TRUE) |>
  dplyr::ungroup() |>
  dplyr::group_by(value) |>
  dplyr::summarise(
    n_tournaments = dplyr::n(),
    .groups = "drop") |>
  dplyr::filter(n_tournaments >= 3)

# Filter ball-by-ball to boundary-scoring balls
boundary_scoring <- t20wc_bbb_tidy |>
  dplyr::filter(runs_off_bat %in% c(4, 6)) |>
  # Filter to teams that played in 3+ tournaments
  dplyr::filter(batting_team %in% teams_in_3_plus_tournaments$value)

```

Compare teams that competed in at least 3 tournaments

- Runs scored from boundaries
- % of innings total from boundaries

## Boundary scoring: Runs scored

```{r wrangle_boundary_scoring_runs}

boundary_scoring_runs <- boundary_scoring |>
  dplyr::select(
    match_id, year, winner, innings, batting_team, dplyr::ends_with("_total"), 
    striker, runs_off_bat) |>
  dplyr::mutate(
    innings_total = dplyr::case_when(
      innings == 1 ~ innings1_total,
      innings == 2 ~ innings2_total),
    innings_label = dplyr::case_when(
      innings == 1 ~ "Setting",
      innings == 2 ~ "Chasing"),
    result_for_batting_team = dplyr::case_when(
      winner == batting_team ~ 1,
      TRUE                   ~ 0),
    innings_label = factor(
      innings_label,
      levels = c("Setting", "Chasing")),
    result_label = dplyr::case_when(
      result_for_batting_team == 1 ~ "Win",
      result_for_batting_team == 0 ~ "Loss"),
    result_label = factor(
      result_label, levels = c("Loss", "Win")),
    batting_team = factor(
      batting_team,
      levels = c("Australia",
                 "England",
                 "India",
                 "New Zealand",
                 "South Africa",
                 "West Indies",
                 "Sri Lanka",
                 "Pakistan",
                 "Bangladesh"))) |>
  dplyr::group_by(
    match_id, year, batting_team, result_label, innings_label, innings_total) |>
  dplyr::summarise(
    count_boundaries = length(runs_off_bat),
    runs_from_boundaries = sum(runs_off_bat),
    .groups = "drop") |>
  dplyr::mutate(
    pct_of_innings_total = round(
      runs_from_boundaries / innings_total * 100, 0)
  )

```

```{r plot_boundary_scoring_runs}

# Build plot
plot_boundary_scoring_runs <- ggplot2::ggplot() +
  ggplot2::geom_boxplot(
    data = boundary_scoring_runs,
    ggplot2::aes(
      x = runs_from_boundaries, y = result_label)) +
  ggplot2::facet_wrap(~batting_team)

# Display plot
plot_boundary_scoring_runs

```

## Boundary scoring: Runs scored as % of innings total

```{r plot_boundary_scoring_pct_of_innings}

# Build plot
plot_boundary_scoring_pct_of_innings <- ggplot2::ggplot() +
  ggplot2::geom_boxplot(
    data = boundary_scoring_runs,
    ggplot2::aes(
      x = pct_of_innings_total, y = result_label)) +
  ggplot2::facet_wrap(~batting_team)

# Display plot
plot_boundary_scoring_pct_of_innings

```

## Building scoreboard pressure

Turning over the strike - balls scored from as % of balls faced (excluding extras)

```{r wrangle_turning_over_strike}

turning_over_strike <- t20wc_bbb_tidy |>
  # Filter to teams that played in 3+ tournaments
  dplyr::filter(batting_team %in% teams_in_3_plus_tournaments$value) |>
  dplyr::mutate(
    flag_runs_off_bat = dplyr::case_when(
      runs_off_bat > 0 ~ 1,
      TRUE             ~ 0)) |>
  dplyr::group_by(year, match_id, batting_team) |>
  dplyr::summarise(
    balls_scored_from = sum(flag_runs_off_bat),
    balls_faced = length(match_id),
    .groups = "drop") |>
  dplyr::mutate(
    pct_of_balls_faced = round(
      balls_scored_from / balls_faced * 100, 0),
    batting_team = factor(
      batting_team,
      levels = c("Australia",
                 "England",
                 "India",
                 "New Zealand",
                 "South Africa",
                 "West Indies",
                 "Sri Lanka",
                 "Pakistan",
                 "Bangladesh")))

```

```{r plot_turning_over_strike}

# Build plot
plot_turning_over_strike <- ggplot2::ggplot() +
  ggplot2::geom_boxplot(
    data = turning_over_strike,
    ggplot2::aes(
      x = year, y = pct_of_balls_faced)) +
  ggplot2::facet_wrap(~batting_team)

# Display plot
plot_turning_over_strike

```

The more successful teams are capable of consistently scoring off ~60% of balls faced (not including runs from extras).

# Bowling performance

## Top order wickets - first 3

How quickly are the top order batters removed?

> "The most important bowling characteristic relate to how quickly the first 3 to 5 batters are dismissed." ([Bhardwaj & Dwyer, 2022](https://doi.org/10.1080/24748668.2022.2045820))

```{r wrangle_wicket_focus}

wicket_focus <- t20wc_bbb_tidy |>
  # Filter to teams that played in 3+ tournaments
  dplyr::filter(bowling_team %in% teams_in_3_plus_tournaments$value) |>
  # Count balls bowled
  dplyr::group_by(match_id, bowling_team) |>
  dplyr::mutate(
    counting_var = 1,
    balls_bowled = cumsum(counting_var)) |>
  dplyr::ungroup() |>
  dplyr::select(
    match_id, year, winner, bowling_team, innings, over, ball, balls_bowled,
    bowler, wicket, wickets_lost_yet, wicket_type, player_dismissed) |>
  dplyr::mutate(
    bowling_team = factor(
      bowling_team,
      levels = rev(c("Australia",
                 "England",
                 "India",
                 "New Zealand",
                 "South Africa",
                 "West Indies",
                 "Sri Lanka",
                 "Pakistan",
                 "Bangladesh"))))

```

```{r wrangle_top_3_wickets}

top_3_wickets <- wicket_focus |>
  dplyr::filter(wickets_lost_yet == 3 & player_dismissed != "")

```

```{r plot_top_3_wickets}

# Build plot
plot_top_3_wickets <- ggplot2::ggplot() +
  ggplot2::geom_boxplot(
    data = top_3_wickets,
    ggplot2::aes(
      x = balls_bowled, y = bowling_team))

# Display plot
plot_top_3_wickets

```

## Dot ball frequency

```{r wrangle_freq_dot_balls}

freq_dot_balls <- t20wc_bbb_tidy |>
  # Filter to teams that played in 3+ tournaments
  dplyr::filter(bowling_team %in% teams_in_3_plus_tournaments$value) |>
  # Count balls bowled
  dplyr::group_by(match_id, bowling_team) |>
  dplyr::mutate(
    counting_var = 1,
    balls_bowled = cumsum(counting_var),
    # Identify dot balls - no runs scored by any method
    flag_dot_balls = dplyr::case_when(
      runs_off_bat == 0 & extras == 0 ~ 1,
      TRUE                            ~ 0)) |>
  # Count dot balls
  dplyr::summarise(
    balls_bowled = max(balls_bowled),
    dot_balls_count = sum(flag_dot_balls),
    .groups = "drop") |>
  dplyr::mutate(
    pct_of_balls_bowled = round(
      dot_balls_count / balls_bowled * 100, 0),
    bowling_team = factor(
      bowling_team,
      levels = rev(c("Australia",
                 "England",
                 "India",
                 "New Zealand",
                 "South Africa",
                 "West Indies",
                 "Sri Lanka",
                 "Pakistan",
                 "Bangladesh"))))

```

```{r plot_freq_dot_balls}

# Build plot
plot_freq_dot_balls <- ggplot2::ggplot() +
  ggplot2::geom_boxplot(
    data = freq_dot_balls,
    ggplot2::aes(
      x = pct_of_balls_bowled, y = bowling_team))

# Display plot
plot_freq_dot_balls

```

## Performance modelling

Logistic regression - W/L

- Runs scored from boundaries
- % of innings total scored from boundaries
- Balls scored from as % of balls faced (excluding extras)
- Balls taken to dismiss the top 3 batters
- Dot balls as % of balls bowled

## Gaps in data

Fielding data availability and range of metrics
- Run outs
- "Dropped" catches? (flagged in Sportradar data)
- Runs saved?

Hawkeye data limited for women's matches

## Extending the analysis

- Strategic / campaign insights - tour timing and frequency, scheduling, familiarity with venue and surfaces, personnel changes (coaches / staff)
- Team composition / selection - international experience, player types (e.g., batting / bowling specialists, types of bowlers - pace / spin, all-rounders), playing list continuity
- Deeper analysis of:
     - Key opposition, using data from T20 internationals in addition to T20 World Cups / Comm Games (e.g., England, India, South Africa, New Zealand)
     - Individual-level performances, identifying impact players with bat and ball, bowler-batter match-ups
- Analysis of performance in game phases - e.g., Powerplay, middle overs, "death" overs
- Data from Sportradar and possibly Hawkeye data if available for women's matches, enabling more granular analysis of:
     - Bowling execution (pitch map, types of balls bowled)
     - Distribution of scoring shots in different parts of the ground
- Relevance of professional T20 leagues? Australia (WBBL), England (The Hundred), India (WPL), New Zealand (Super Smash)
