---
title: "What It Takes To Win: Cricket - T20"
author: "Jacquie Tran"
format: revealjs
---

## Introduction

- Cricket will be added to the Olympic sports program for LA 2028. The format is expected to be T20.
- The Australian Cricket Teams are consistently among the most successful international sides in world cricket, the Women's Team in particular.
- In recent years, with the professionalisation and maturation of Women's Cricket globally, more nations are challenging the Australian Women's Team in major events (see: T20 World Cup 2024, won by the White Ferns).
- Timely and important to understand What It Takes To Win in Women's T20 Cricket, at the start of the 2028 Olympic cycle.

## Only a slice of "What It Takes To Win"

- Using publicly available game stats
- Acknowledgement that other factors are critical (daily training and recovery measures, trusted coaches and support team, physiological / biomechanical measures, etc.) but not reflected in public data and thus not able to be analysed for this presentation

## Higher-level principles of team sports performance

Mix of experience, early / mid / late career

Adaptability: in tournament play, in ways to impact the game (all-rounders)

Balanced performance across all major facets of the game (bat, ball, fielding)

Unpredictability (options as a batter, left-right batting partnerships, variety as a bowler, uncommon bowling styles - leggie, left arm orthodox)

Impact play: Batting - Boundary scoring rate, strike rate. Bowling - Strike rate, wickets in key phases (early innings Power Play, late innings / "death" overs?)

## Motivating questions / hypotheses

Descriptive analysis: Progression of standard metrics over multiple tournaments

- Innings total when setting vs. chasing (wins and losses)
- Wickets taken
- Batting averages
- Strike rates
- Boundary scoring rate
- Extras bowled
- Modes of dismissal
- Spread of bowlers used

Team make-up

- Depth chart: Age distribution, international experience, World Cup experience / tournament continuity
- All-rounders: Number selected and output with bat and ball
- Stability of playing XI

Performance

- Batting anchor: "Build around Beth Mooney"
- Building pressure with the bat:
     - Capability to accelerate strike rate
     - Turning over the strike
     - Distribution of shots in different parts of the ground
- Building pressure with the ball:
     - Dot balls, wickets, partnership bowling, variety
     - Fielding / modes of dismissal
- Variety of contributors with bat and ball (minimising dependence on "star" players)

## Methods

- This presentation will focus on Women's T20 World Cups and Commonwealth Games (since 2028 will be the first appearance at an Olympic Games for cricket)
- Data obtained from [Cricsheet.org](https://cricsheet.org) using the `{cricketdata}` R package
- Ball-by-ball data available for T20 World Cups in 2018, 2020, 2023, and 2024, and the 2022 Commonwealth Games

## Retrieve data

```{r retrieve_data}

# First, obtain match IDs for Women's T20 World Cup tournaments or Comm Games
t20wc_match_ids <- cricketdata::fetch_cricsheet(
  type = "match",
  gender = "female",
  competition = "t20s") |>
  # Filter to World Cup matches
  dplyr::filter(
    event %in% c(
      "ICC Women's T20 World Cup",
      "Commonwealth Games Women's Cricket Competition"))

# Retrieve playing XI lists, filter to T20 World Cup or Comm Games matches only
t20wc_team_lists <- cricketdata::fetch_cricsheet(
  type = "player",
  gender = "female",
  competition = "t20s") |>
  # Filter to World Cup matches
  dplyr::filter(
    match_id %in% t20wc_match_ids$match_id)

# Retrieve ball-by-ball data, filter to T20 World Cup or Comm Games matches only
t20wc_bbb <- cricketdata::fetch_cricsheet(
  type = "bbb",
  gender = "female",
  competition = "t20s") |>
  # Filter to World Cup matches
  dplyr::filter(
    match_id %in% t20wc_match_ids$match_id)

```

**JT NOTE:** In team lists data, do I have to tidy up any name changes over time?

```{r}

# Retrieve career data from ESPNCricinfo
EllysePerry <- cricketdata::fetch_player_data(275487, "T20", "batting")

```


## Gaps

Fielding data availability and range of metrics

Hawkeye data limited for women's matches

## Extending the analysis

- Deeper analysis of key opposition, using data from T20 internationals in addition to T20 World Cups / Comm Games (e.g., England, India, South Africa, New Zealand)
- Data from Sportradar and possibly Hawkeye data if available for women's matches
- Relevance of professional T20 leagues? Australia (WBBL), England (The Hundred), India (WPL), New Zealand (Super Smash)
